---
title: "Assignment 5: Exercise in Data Visualization."
author: "Oscar Feng-Hsun Chang"
output: html_document
---

In this assignment, I used baseball statistic data downloaded from [FanGraphs website](http://www.fangraphs.com) simply because I'm interested in baseball. This data include the following statistics of all the starting pitchers in the past 30 seasons (1986-2016). 

+ W: game winned
+ L: game lost
+ G: game played
+ GS: game played as the starter
+ IP: inning played in the whole season
+ K9: strike out produced per 9 innings
+ BB9: Balls produced per 9 innings
+ BABIP: how many of a batter's balls in play go for hits,  excluding home runs
+ LOB: Left on Base percentage,  the percentage of base runners that a pitcher strands on base over the course of a season
+ HRFB: Home Run to Fly Ball rate, ratio of how many home runs are hit against a pitcher for every fly ball they allow
+ FIP: Fielding Independent Pitching, what a player¡¦s ERA would look like over a given period of time if the pitcher were to have experienced league average results on balls in play and league average timing
+ WAR: Wins Above Replacement, an attempt by the sabermetric baseball community to summarize a player¡¦s total contributions to their team in one statistic
+ WHIP: Walks plus Hits per Innings Pitched, measures how many base runners a pitcher lets up
+ ERA: Earned Run Average, calculated by dividing a pitcher¡¦s total number of Earned Runs allowed by his total number of Innings Pitched and multiply by nine
+ Age: the pitcher's age at the season


### Loading packages and data

First, I loaded the packages that are needed in my further analyses. Note that I use dplyr (I think dplyr is a special version of plyr that is designed specifically for data frame) and tidyr instead of plyr and reshape2. However, in order to demonstrate that I can manage reshape2 package, I duplicate some of my code in two packages. 

```{r, load package, message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(plyr)
library(reshape2)
```

Now, load the data and skim the data type and data structure.

```{r,load data}
setwd("D:/Courses/UM/2016_WN/EEB480_modelinfer/20160926")
AL_SP = read.csv(file="MLB_AL_SP_1986_2016.csv", header = TRUE, sep=",", fill=TRUE)
str(AL_SP)
head(AL_SP)
```

Now use _arrange_ function to sort the data first by Season(year) then by team and Pitcher's name, so that it looks nicer.

```{r, order data}
AL_SP = AL_SP %>% arrange(Season, Team, Name)
head(AL_SP)
```

### Answering couple easy questions


1.In the past 30 seasons, how many starting pitchers pitched at least 1 inning and 1 game? I also make these pitchers into a new dataframe (ALSP.all). Note that I used _filter_ in _dplyr_ package, but I can also use _subset_ in _plyr_ package. 

```{r, filter and subset}
ALSP.all = AL_SP %>%
           filter(IP>1 & GS>1)
ALSP.all2 = AL_SP %>%
           subset(IP>1 & GS>1)

allSP = length(unique(ALSP.all$Name))
allSP
```

2.What are the average and standard deviation of each statistics of each year? To answer this question, I first have to reshape the data into a "long" form, so I use _gather_ in _tidyr_ package and _melt_ in _reshape2_ package. 

```{r, long to wide}
ALSP.all.long = ALSP.all %>%
                gather(key=stats, value=val, W, L, G, GS, IP, K9, BB9, HR9, BABIP, LOB, HRFB, FIP, WAR, WHIP, ERA, Age)
ALSP.all.long2 = ALSP.all %>%
                 melt(id.vars=colnames(ALSP.all)[1:3], variable.name="stats", value.name="val")
# melt seems to be more convenient as it can take the output from "colnames" directly

ALSP.all.long2$stats=as.character(ALSP.all.long2$stats) 
# Beacuse melt function makes the variables (names as stats here) into factor, so I'll have to transform it 
# into character for comparing the two dataframe.  
all_equal(target=ALSP.all.long, current=ALSP.all.long2)
```

Just for fun, I can _cast_ (in _reshape2_) or _spread_ (in _tidyr_) the long data frame back to wide form and it should be identical to the "ALSP.all"

```{r, wide to long}
ALSP.all.wide = ALSP.all.long %>% 
                spread(key=stats, value=val)

ALSP.all.wide2 = ALSP.all.long2 %>%
                 dcast(Season+Team+Name~stats, value.var=c("val"))

all_equal(target=ALSP.all.wide, current=ALSP.all.wide2)
all_equal(target=ALSP.all.wide, current=ALSP.all, convert=TRUE)
all_equal(target=ALSP.all.wide2, current=ALSP.all, convert=TRUE)
```

With the long form dataframe, I can use _ddply_ function to calculate the mean and standard deviation for each statistics. But it will be very long, so I just show the first 6 rows of this summary dataframe.

```{r, summarizing data}
SeasonSum = ddply(ALSP.all.long, ~stats+Season, summarize, mean=mean(val), sd=sd(val))
head(SeasonSum)
```

I then visualize this summary dataframe by histogram (a very busy one) and several boxplots.

```{r, visulaizing summary dataframe_W/L, fig.width=10}
ggplot(data=ALSP.all.long[which(ALSP.all.long$stats=="W"|ALSP.all.long$stats=="L"),], 
       aes(x=factor(Season), y=val, color=stats),
       group=stats)+
  geom_boxplot(position="dodge")+ #self note, postition argument by default is "dodge"
  theme(axis.text.x=element_text(angle = 90, hjust=1))+
  xlab("Season")+
  ylab("value")
```

Interestingly, on average, a pitcher tend to loss more than win a game.

```{r, visulaizing summary dataframe, fig.width=10}
ggplot(data=ALSP.all.long[which(ALSP.all.long$stats=="G"|ALSP.all.long$stats=="GS"),], 
       aes(x=factor(Season), y=val, color=stats),
       group=stats)+
  geom_boxplot(position=position_dodge(width=0.8))+ #self note, or I can set how much each bar should dodge manually
  theme(axis.text.x=element_text(angle = 90, hjust=1))+
  xlab("Season")+
  ylab("value")

ggplot(data=ALSP.all.long[which(ALSP.all.long$stats=="K9"|ALSP.all.long$stats=="BB9"|ALSP.all.long$stats=="HR9"),], 
       aes(x=factor(Season), y=val, color=stats),
       group=stats)+
  geom_boxplot(position="dodge")+
  facet_grid(stats~., scales="free")+
  theme(axis.text.x=element_text(angle = 90, hjust=1))+
  xlab("Season")+
  ylab("value")

ggplot(data=ALSP.all.long[which(ALSP.all.long$stats=="IP"|ALSP.all.long$stats=="Age"),], 
       aes(x=factor(Season), y=val, color=stats),
       group=stats)+
  geom_boxplot(position="dodge")+
  facet_grid(stats~., scales="free")+
  theme(axis.text.x=element_text(angle = 90, hjust=1))+
  xlab("Season")+
  ylab("value")

ggplot(data=ALSP.all.long[which(ALSP.all.long$stats=="ERA"|ALSP.all.long$stats=="FIP"),], 
       aes(x=factor(Season), y=val, group=interaction(Season, stats), color=stats))+
  geom_boxplot(position="dodge")+
  facet_grid(stats~., scales="free")+
  scale_y_continuous(limits = c(0, 10))+
  theme(axis.text.x=element_text(angle = 90, hjust=1))+
  xlab("Season")+
  ylab("value")

ggplot(data=ALSP.all.long[which(ALSP.all.long$stats=="BABIP"|ALSP.all.long$stats=="WHIP"|ALSP.all.long$stats=="WAR"),], 
       aes(x=factor(Season), y=val, color=stats),
       group=stats)+
  geom_boxplot(position="dodge")+
  facet_grid(stats~., scales="free")+
  theme(axis.text.x=element_text(angle = 90, hjust=1))+
  xlab("Season")+
  ylab("value")

ggplot(data=ALSP.all.long[which(ALSP.all.long$stats=="LOB" | ALSP.all.long$stats=="HRFB"),], 
       aes(x=factor(Season), y=val, color=stats),
       group=stats)+
  geom_boxplot(position="dodge")+
  facet_grid(stats~., scales="free")+
  theme(axis.text.x=element_text(angle = 90, hjust=1))+
  xlab("Season")+
  ylab("value")
# There will be a warning message as HRFB is not available prior to 2001.
```

3.It seems there is not much change of the average of each statistics over the past 30 seasons. How about pitchers who pitched more than 50IP and 100IP?
```{r, selecting 50IP and 100IP}
ALSP.50IP = AL_SP %>%
           filter(IP>50)
ALSP.100IP = AL_SP %>%
           subset(IP>100)
ALSP.50IP.long = ALSP.50IP %>%
                gather(key=stats, value=val, W, L, G, GS, IP, K9, BB9, HR9, BABIP, LOB, HRFB, FIP, WAR, WHIP, ERA, Age)
ALSP.100IP.long = ALSP.100IP %>%
                gather(key=stats, value=val, W, L, G, GS, IP, K9, BB9, HR9, BABIP, LOB, HRFB, FIP, WAR, WHIP, ERA, Age)
```

```{r, visualizing summary data_50IP and 100IP, fig.width=10}
ggplot(data=ALSP.50IP.long[which(ALSP.50IP.long$stats=="W"|ALSP.50IP.long$stats=="L"),], 
       aes(x=factor(Season), y=val, color=stats),
       group=stats)+
  geom_boxplot(position="dodge")+ #self note, postition argument by default is "dodge"
  theme(axis.text.x=element_text(angle = 90, hjust=1))+
  xlab("Season")+
  ylab("value")

ggplot(data=ALSP.100IP.long[which(ALSP.100IP.long$stats=="W"|ALSP.100IP.long$stats=="L"),], 
       aes(x=factor(Season), y=val, color=stats),
       group=stats)+
  geom_boxplot(position="dodge")+ #self note, postition argument by default is "dodge"
  theme(axis.text.x=element_text(angle = 90, hjust=1))+
  xlab("Season")+
  ylab("value")
```

3.1 I'm then interested to see if a pitcher pitched more innings, can that pitcher winn more games?

```{r}
ALSP.50IP.long = ALSP.50IP %>%
                 mutate(WL = W/L) %>%
                 melt(id.vars=colnames(ALSP.50IP)[1:3], variable.name="stats", value.name="val")

ALSP.100IP.long = ALSP.100IP %>%
                  mutate(WL = W/L) %>%
                  melt(id.vars=colnames(ALSP.100IP)[1:3], variable.name="stats", value.name="val")

```


```{r, visulaizing summary dataframe, fig.width=10}
ggplot(data=ALSP.all.long[which(ALSP.all.long$stats=="G"|ALSP.all.long$stats=="GS"),], 
       aes(x=factor(Season), y=val, color=stats),
       group=stats)+
  geom_boxplot(position=position_dodge(width=0.8))+ #self note, or I can set how much each bar should dodge manually
  theme(axis.text.x=element_text(angle = 90, hjust=1))+
  xlab("Season")+
  ylab("value")

ggplot(data=ALSP.all.long[which(ALSP.all.long$stats=="K9"|ALSP.all.long$stats=="BB9"|ALSP.all.long$stats=="HR9"),], 
       aes(x=factor(Season), y=val, color=stats),
       group=stats)+
  geom_boxplot(position="dodge")+
  facet_grid(stats~., scales="free")+
  theme(axis.text.x=element_text(angle = 90, hjust=1))+
  xlab("Season")+
  ylab("value")

ggplot(data=ALSP.all.long[which(ALSP.all.long$stats=="IP"|ALSP.all.long$stats=="Age"),], 
       aes(x=factor(Season), y=val, color=stats),
       group=stats)+
  geom_boxplot(position="dodge")+
  facet_grid(stats~., scales="free")+
  theme(axis.text.x=element_text(angle = 90, hjust=1))+
  xlab("Season")+
  ylab("value")

ggplot(data=ALSP.all.long[which(ALSP.all.long$stats=="ERA"|ALSP.all.long$stats=="FIP"),], 
       aes(x=factor(Season), y=val, group=interaction(Season, stats), color=stats))+
  geom_boxplot(position="dodge")+
  facet_grid(stats~., scales="free")+
  scale_y_continuous(limits = c(0, 10))+
  theme(axis.text.x=element_text(angle = 90, hjust=1))+
  xlab("Season")+
  ylab("value")

ggplot(data=ALSP.all.long[which(ALSP.all.long$stats=="BABIP"|ALSP.all.long$stats=="WHIP"|ALSP.all.long$stats=="WAR"),], 
       aes(x=factor(Season), y=val, color=stats),
       group=stats)+
  geom_boxplot(position="dodge")+
  facet_grid(stats~., scales="free")+
  theme(axis.text.x=element_text(angle = 90, hjust=1))+
  xlab("Season")+
  ylab("value")

ggplot(data=ALSP.all.long[which(ALSP.all.long$stats=="LOB" | ALSP.all.long$stats=="HRFB"),], 
       aes(x=factor(Season), y=val, color=stats),
       group=stats)+
  geom_boxplot(position="dodge")+
  facet_grid(stats~., scales="free")+
  theme(axis.text.x=element_text(angle = 90, hjust=1))+
  xlab("Season")+
  ylab("value")
# There will be a warning message as HRFB is not available prior to 2001.

```


4. 














