---
title: "Revised Assignment 5: Exercise in Data Visualization_20161019"
author: "Oscar Feng-Hsun Chang"
output: html_document
---

In this assignment, I used baseball statistic data downloaded from [FanGraphs website](http://www.fangraphs.com) simply because I'm interested in baseball. This data include the following statistics of all the starting pitchers in the past 30 seasons (1986-2016).

+ W: game won
+ L: game lost
+ G: game played
+ GS: game played as the starter
+ IP: inning played in the whole season
+ K9: strike out produced per 9 innings
+ BB9: Balls produced per 9 innings
+ BABIP: how many of a batter's balls in play go for hits, excluding home runs
+ LOB: Left on Base percentage,  the percentage of base runners that a pitcher strands on base over the course of a season
+ HRFB: Home Run to Fly Ball rate, ratio of how many home runs are hit against a pitcher for every fly ball they allow
+ FIP: Fielding Independent Pitching, what a player??s ERA would look like over a given period of time if the pitcher were to have experienced league average results on balls in play and league average timing
+ WAR: Wins Above Replacement, an attempt by the sabermetric baseball community to summarize a player's total contributions to their team in one statistic
+ WHIP: Walks plus Hits per Innings Pitched, measures how many base runners a pitcher lets up
+ ERA: Earned Run Average, calculated by dividing a pitcher??s total number of Earned Runs allowed by his total number of Innings Pitched and multiply by nine
+ Age: the pitcher's age at the season


### Loading packages and data

First, I loaded the packages that are needed in my further analyses. 

```{r, load package, message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(plyr)
library(reshape2)
library(RCurl)
```

Now, load the data.

```{r,load data}
AL_SP=read.csv(text=getURL("https://raw.githubusercontent.com/OscarFHC/EEB480_ModelInfer_assignment/master/Mod_Infer_hw5_EDA/MLB_AL_SP_1986_2016.csv"), sep=",", header=T, comment.char="#")
```

Now use _arrange_ function to sort the data first by Season(year) then by team and Pitcher's name, so that it looks nicer. After that I filter out pitchers that pitched more than 1 innings. 

```{r, order data}
AL_SP.all = AL_SP %>% 
  arrange(Season, Team, Name) %>%
  filter(IP>1 & GS>1)
head(AL_SP.all)
```

## 1.What is the relationship between ERA (Earned Run Average) and the Win-loss ratio and does this relationship change with IP (number of Inning Pitched).

To make the win-loss ratio comparable, I divided the ratio by the IP. I also divide the pitcher's into 3 arbitrary groups according to the IP (<50, 50-100, >100)

```{r, echo=FALSE}
small50IP = AL_SP %>%
            filter(IP<50, W>0, L>0) %>%
            mutate(WL9 = W/L/IP, IP.id="<50 IP")

large50IP = AL_SP %>%
            filter(IP>50 & IP<100, W>0, L>0) %>%
            mutate(WL9 = W/L/IP, IP.id="50-100 IP")

large100IP = AL_SP %>%
             filter(IP>100, W>0, L>0) %>%
             mutate(WL9 = W/L/IP, IP.id=">100IP")

ALSP.WL = rbind(small50IP, large50IP, large100IP) %>%
          filter(WL9>=0)%>%
          select(Season, Name, IP, IP.id, WL9, ERA, WAR, BABIP)
```

```{r, echo=FALSE}
ggplot(data=ALSP.WL,
       group=IP.id)+
  geom_point(aes(x=ERA, y=WL9))+
  #facet_grid(IP.id~., scales="free")+ # this is not working
  facet_wrap(~IP.id, nrow=1, scales="free")+
  xlab("ERA")+
  ylab("Win-Loss ratio per 9 inning pitched")
```

To me, this reveals two things. First, giving up more runs does not necessary means losing more, which could make sense as winning a game requires not just allowing less runs but also batting in more runs. Second, the relationship between Win-Loss ratio and ERA, if there is any, looks non-linear, which is even more interesting!

I then try to plot a 3D plot (which basically modifies the code you wrote) to visualize the relationship between IP, ERA and Win-Loss ratio per 9 inning pitched. 

```{r }
ERA.breaks = with(ALSP.WL, seq(from=min(ERA),to=max(ERA), length=51))
IP.breaks = with(ALSP.WL, seq(from=min(IP),to=max(IP), length=51))
ALSP.WL %>%
  mutate(
    ERA.bin=cut(ERA, breaks=ERA.breaks, include=TRUE,ordered=TRUE),
    IP.bin=cut(IP, breaks=IP.breaks, include=TRUE,ordered=TRUE)
  ) %>% 
  mutate(
    ERA=mapvalues(as.numeric(ERA.bin), from=seq_len(50), to=head(ERA.breaks,-1)+0.5*diff(ERA.breaks)),
    IP=mapvalues(as.numeric(IP.bin), from=seq_len(50), to=head(IP.breaks,-1)+0.5*diff(IP.breaks))
    ) %>%
  ddply(~ERA.bin+IP.bin,summarize,
        ERA=mean(ERA),WL9=mean(WL9),IP=mean(IP)) %>% 
  ggplot(aes(x=ERA, y=IP, z=WL9))+
  geom_tile(aes(fill=WL9))+
  xlab("ERA")+
  ylab("IP")
```

Here, it seems that at similar ERA, the more inning a pitcher pitched, the less likely the pitcher produce more win. This is somehow counter intuitive to me. I suspect this is just a coincident. 

## 2. What is the relationship between BABIP (should reflect how many hit a pitcher allowed per balls in play) and ERA (Earned Run Average) and how does this relationship change with IP.

```{r}
ggplot(data=ALSP.WL, aes(x=ERA, y=BABIP),
       group=IP.id)+
    geom_point()+
    stat_smooth(method="lm")+
    facet_wrap(~IP.id, nrow=1, scales="free")+
    xlab("BABIP")+
    ylab("ERA")
```

There is a more clear positive relationship between BABIP and ERA, indicating that the more hit a pitcher allowed, the more runs the pitcher will allow. This relationship also seems to be different in each group of pitcher. Again, I can use contour plot to avoid arbitrary grouping.
