---
title: "Revised Assignment 5: Exercise in Data Visualization_20161019"
author: "Oscar Feng-Hsun Chang"
output: html_document
---

In this assignment, I used baseball statistic data downloaded from [FanGraphs website](http://www.fangraphs.com) simply because I'm interested in baseball. This data include the following statistics of all the starting pitchers in the past 30 seasons (1986-2016).

+ W: game won
+ L: game lost
+ G: game played
+ GS: game played as the starter
+ IP: inning played in the whole season
+ K9: strike out produced per 9 innings
+ BB9: Balls produced per 9 innings
+ BABIP: how many of a batter's balls in play go for hits, excluding home runs
+ LOB: Left on Base percentage,  the percentage of base runners that a pitcher strands on base over the course of a season
+ HRFB: Home Run to Fly Ball rate, ratio of how many home runs are hit against a pitcher for every fly ball they allow
+ FIP: Fielding Independent Pitching, what a player??s ERA would look like over a given period of time if the pitcher were to have experienced league average results on balls in play and league average timing
+ WAR: Wins Above Replacement, an attempt by the sabermetric baseball community to summarize a player's total contributions to their team in one statistic
+ WHIP: Walks plus Hits per Innings Pitched, measures how many base runners a pitcher lets up
+ ERA: Earned Run Average, calculated by dividing a pitcher??s total number of Earned Runs allowed by his total number of Innings Pitched and multiply by nine
+ Age: the pitcher's age at the season


### Loading packages and data

First, I loaded the packages that are needed in my further analyses. Note that I use dplyr (I think dplyr is a special version of plyr that is designed specifically for data frame) and tidyr instead of plyr and reshape2. However, in order to demonstrate that I can manage reshape2 package, I duplicate some of my code in two packages.

> Just use whichever you prefer: there is no need to do the same thing twice.

```{r, load package, message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(plyr)
library(reshape2)
library(RCurl)
```

Now, load the data and skim the data type and data structure.

```{r,load data}
AL_SP=read.csv(text=getURL("https://raw.githubusercontent.com/OscarFHC/EEB480_ModelInfer_assignment/master/Mod_Infer_hw5_EDA/MLB_AL_SP_1986_2016.csv"), sep=",", header=T, comment.char="#")

str(AL_SP)
head(AL_SP)
```

> Don't show this to your reader unless you want them to see it. If you want a record of these procedures, hide them from the reader using chunk option `include=FALSE`.

Now use _arrange_ function to sort the data first by Season(year) then by team and Pitcher's name, so that it looks nicer.

```{r, order data}
AL_SP = AL_SP %>% arrange(Season, Team, Name)
head(AL_SP)
```

### Answering couple easy questions

## 1.In the past 30 seasons, how many starting pitchers pitched at least 1 inning and 1 game? I also make these pitchers into a new dataframe (ALSP.all). Note that I used _filter_ in _dplyr_ package, but I can also use _subset_ in _plyr_ package.

```{r, filter and subset}
ALSP.all = AL_SP %>%
           filter(IP>1 & GS>1)
ALSP.all2 = AL_SP %>%
           subset(IP>1 & GS>1)

allSP = length(unique(ALSP.all$Name))
allSP
```

> Comments? What do you compare this figure to?

## 2.Does a pitcher win or lose more games across the past 30 seasons?

```{r}
ALSP.all.long = ALSP.all %>%
                gather(key=stats, value=val, W, L, G, GS, IP, K9, BB9, HR9, BABIP, LOB, HRFB, FIP, WAR, WHIP, ERA, Age)

ggplot(data=ALSP.all.long[which(ALSP.all.long$stats=="W"|ALSP.all.long$stats=="L"),],
       aes(x=factor(Season), y=val, color=stats),
       group=stats)+
  geom_boxplot(position="dodge")+ #self note, postition argument by default is "dodge"
  theme(axis.text.x=element_text(angle = 90, hjust=1))+
  xlab("Season")+
  ylab("number of games a pitcher Wins or Loses")

```

> What does this plot tell us? Why do we care about the distribution of wins/losses for a given year?

Supprisingly, on average, a pitcher tend to loss more than win a game, although the differences are not significant. This is probably due to the statistical artifact of this stats that losing a game is a little bit easier than winning a game. To be recorded as a losing pitcher, a pitcher only needs to allow a run which gives the opposing team the lead regardless how many innings that pitcher pitched. But to be recorded as a winning pitcher, a pitcher has to pitch at least 5 innings if the pitcher is the starting pitcher.  

## 3.What is the relationship between ERA (Earned Run Average) and the Win-loss ratio and does this relationship change with IP (number of Inning Pitched).

To make the win-loss ratio comparable, I divided the ratio by the IP. I also divide the pitcher's into 3 arbitrary groups according to the IP (<50, 50-100, >100)

```{r, echo=FALSE}
small50IP = AL_SP %>%
            filter(IP<50, W>0, L>0) %>%
            mutate(WL9 = W/L/IP, IP.id="<50 IP")

large50IP = AL_SP %>%
            filter(IP>50 & IP<100, W>0, L>0) %>%
            mutate(WL9 = W/L/IP, IP.id="50-100 IP")

large100IP = AL_SP %>%
             filter(IP>100, W>0, L>0) %>%
             mutate(WL9 = W/L/IP, IP.id=">100IP")

ALSP.WL = rbind(small50IP, large50IP, large100IP) %>%
          filter(WL9>=0)%>%
          select(Season, Name, IP, IP.id, WL9, ERA, WAR, BABIP)
```

```{r, echo=FALSE}
ggplot(data=ALSP.WL,
       group=IP.id)+
  geom_point(aes(x=ERA, y=WL9))+
  #facet_grid(IP.id~., scales="free")+ # this is not working
  facet_wrap(~IP.id, nrow=1, scales="free")+
  xlab("ERA")+
  ylab("Win-Loss ratio per 9 inning pitched")
```

To me, this reveals two things. First, giving up more runs does not necessary means losing more, which could make sense as winning a game requires not just allowing less runs but also batting in more runs. Second, the relationship between Win-Loss ratio and ERA, if there is any, looks non-linear, which is even more interesting!

---------------------------------------------------------------------------------------------------------------------------

* Couple questions here.
1. The scale of x-axis is fixed even I used the "scales="free" " argument in the `facet_grid()` function, so I have to use `facet_wrap()`.  

> Yes, but is that a question?

2. If I'm interested to see how the relationship change with IP, I guess I should do a 3D plot or a contour instead of a arbitrary grouping. The following code gives me warning message saying
"Computation failed in `stat_contour()`:
dimensions of 'x', 'y' and 'z' do not match "
But the dimension of the x, y, and z are the same. Please see the following code.

> The data need to be in a grid format for `geom_contour`. Perhaps try binning the data?

```{r eval=FALSE}
ggplot(data=ALSP.WL, aes(x=ERA, y=WL9, z=IP))+
  geom_contour(binwidth = 0.001)+
  xlab("ERA")+
  ylab("Win-Loss ratio per 9 inning pitched")
```

```{r }
ERA.breaks <- with(ALSP.WL,seq(from=min(ERA),to=max(ERA),length=51))
WL9.breaks <- with(ALSP.WL,seq(from=min(WL9),to=max(WL9),length=51))
ALSP.WL %>%
  mutate(
    ERA.bin=cut(ERA,breaks=ERA.breaks,include=TRUE,ordered=TRUE),
    WL9.bin=cut(WL9,breaks=WL9.breaks,include=TRUE,ordered=TRUE)
  ) %>% 
  mutate(
    ERA=mapvalues(as.numeric(ERA.bin),from=seq_len(50),
                  to=head(ERA.breaks,-1)+0.5*diff(ERA.breaks)),
    WL9=mapvalues(as.numeric(WL9.bin),from=seq_len(50),
                  to=head(WL9.breaks,-1)+0.5*diff(WL9.breaks))
    ) %>%
  ddply(~ERA.bin+WL9.bin,summarize,
        ERA=mean(ERA),WL9=mean(WL9),IP=mean(IP)) %>% 
  ggplot(aes(x=ERA, y=WL9, z=IP))+
  geom_tile(aes(fill=IP))+
  # geom_contour(binwidth = 0.001)+
  xlab("ERA")+
  ylab("Win-Loss ratio per 9 inning pitched")
```



## 4. What is the relationship between BABIP (should reflect how many hit a pitcher allowed per balls in play) and ERA (Earned Run Average) and how does this relationship change with IP.

```{r}
ggplot(data=ALSP.WL, aes(x=ERA, y=BABIP),
       group=IP.id)+
    geom_point()+
    stat_smooth(method="lm")+
    facet_wrap(~IP.id, nrow=1, scales="free")+
    xlab("BABIP")+
    ylab("ERA")
```

There is a more clear positive relationship between BABIP and ERA, indicating that the more hit a pitcher allowed, the more runs the pitcher will allow. This relationship also seems to be different in each group of pitcher. Again, I can use contour plot to avoid arbitrary grouping.
