---
title: "hw7_FHChang"
author: "Feng-Hsun Chang"
date: "2016.Nov.9"
output: 
  html_document:
    code_folding: show
    highlight: haddock
    number_sections: FALSE
    theme: yeti
    toc: FALSE
    toc_float:
      collapsed: FALSE
      smooth_scroll: TRUE
---

In this assignment, I use for loop to calculate the agreement between observation and predicted probability for different numbers of seed availability.  

To calculat ethe agreement between observation and predicted probability, calculate the sum of square between observation and predicted probability.  

In each loop, I first calculate the moments of both binomial ($n$ and $p$) and beta-binomial distribution($p$ and $\theta$) from the observation. I then used these estimated parameters to predict the probability of number of seeds being predated.  

```{r}
library(magrittr)
library(plyr)
library(ggplot2)
library(emdbook) 
library(reshape2)

seeds = read.csv("http://kinglab.eeb.lsa.umich.edu/480/data/seedpred.csv",
         comment.char="#",
         colClasses=c(date="Date",station='factor',
                      dist='factor',species='factor')) %>%
  arrange(station,dist,species,date)

seeds %>%
  ddply(~station+dist+species,summarize,
        avail=head(seeds,-1),
        taken=-diff(seeds),
        tint=diff(date)) %>%
  subset(avail>0) -> dat

moms.binom.all = c()
moms.bb.all = c()
pdf.all = c()
error.all = c()

for (i in 2:5){
  subseeds <- subset(dat,avail==i)
  pl = ggplot(data=subseeds,mapping=aes(x=taken))+
    geom_histogram(binwidth=1,fill=NA,color='black',
                   mapping=aes(y=..density..))
  
  moms.binom = subseeds %>% summarize(n=unique(avail),p=mean(taken)/n)
    ## calculate the moments of binomial distribution, which are the number of seed (n) the probability of seed predation (p)
  moms.binom.all = rbind(moms.binom.all, moms.binom)
    ## store the moments of binomial distribution to a data frame, "moms.binom.all"
  pdf = data.frame(x=seq(0,i)) %>%
    mutate(binom=dbinom(x=x,prob=moms.binom$p,size=moms.binom$n))
    ## claculate the probability of certain number of seeds being predated according to a binomial distribution with parameters ("n" and "p") estimated from the observation.
  
  moms.bb = subseeds %>% 
    summarize(n=unique(avail),m=mean(taken),v=var(taken),
              p=m/n,theta=-n*(v-m*(n-m))/(n*v-m*(n-m)))
    ## calculate hte moments of beta-binomial distribution, which are the probability (p) and dispersion parameter (theta)
  moms.bb.all = rbind(moms.bb.all, moms.bb)
    ## store the moments of beta-binomial distribution to a data frame, "moms.bb.all"
  pdf = pdf %>% mutate(betabinom=dbetabinom(x=x,size=moms.bb$n,prob=moms.bb$p,theta=moms.bb$theta),
                  obs = count(subseeds, vars = "taken")[,2]/nrow(subseeds))
    ## claculate the probability of certain number of seeds being predated according to a beta-binomial distribution with parameters ("p" nad "theta") estimated from the observation.
    ## Also calculate the observed probability of seed predation
  pdf.l = pdf %>% melt(id.var=c("x"), variable.name="fun", value.name="den")
    ## preparing long data form for figure
  pdf.all = rbind(pdf.all, pdf)
    ## combine all the probabilities together
  
  p.final = pl+
    geom_point(data = pdf.l[which(pdf.l$fun!="obs"),], aes(x=x, y=den, group=fun, color=factor(fun)), size=3)+
    scale_colour_manual(name="distribution \n function",
                        breaks=c("betabinom", "binom"),
                        labels=c("betabinomial", "binomial"),
                        values=c("red","blue"))+
    labs(x="number of seeds taken", y="probability")+
    theme_bw()+
    theme(panel.grid.major=element_blank(),
          panel.grid.minor=element_blank(),
          strip.background = element_blank(),
          panel.border=element_rect(color="black"),
          legend.key = element_blank())
  print(p.final)
  
  error = pdf %>% 
    summarize(ebinom = sqrt(sum((binom-obs)^2)), ebb = sqrt(sum((betabinom-obs)^2)))
    ## Here I calculate the sum of square of observe versus predicted probability as a way to demonstrate the agreement between data and prediction.
  error.all = rbind(error.all, error)
    ## I store the sum of square of the two model distributions, so that in the end I can see how the agreement change with number of seeds available. 
}

```


These plots are the visualization of observation versus predicted probability in different seed number availability. The observation and predicted probability, I use the histogram is the observation, red dots are the prediction from binimial distribution, blue dots are the prediction from beta-binomial distribution. 

```{r}
library(knitr)
error.all = cbind(seq.int(from=2,to=5), error.all)
  colnames(error.all) = c("seed availibility", "error of binomial distribution", "error of beta-binomial distribution")
kable(error.all)
```

From the above table, we see that the sum of square gradually increase with the number of available seeds. This means that the agreement between observation and predicted probability gradually ddecrease with number of comparisons between observation and predicted probability. In addition, the beta-binomial distribution always performs better than the binimial distribution (lower sum of square).  

I'm wondering if I should divide each sum of square by the number of comparisons in each subset of the data. If yes, the sum of square would be as the following table. 

```{r}
error.all = error.all %>% 
  mutate(meanSS_binom = error.all[,"error of binomial distribution"]/(error.all[,"seed availibility"]+1),
         meanSS_bb = error.all[,"error of beta-binomial distribution"]/(error.all[,"seed availibility"]+1))

  colnames(error.all) = c(colnames(error.all)[1:3], "mean error of binomial distribution", "mean error of beta-binomial distribution")

kable(error.all)
```

This table show similar pattern that with the increase of the number of available seeds, the agreement gradually decrease. The beta-binimial distribution also always performs better than the binimial distribution. 


Question: I understand that beta-binomial distribution is the binomial distribution where the probability ($p$) is a random variable following beta distribution (with $\alpha$ and $\beta$ parameters). I still have trouble think through the $\theta$ parameter and the effects of it.  